--- 
title: "TMA4315: Compulsory exercise 3: (Generalized) Linear Mixed Models" 
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
    toc_float: yes
date: "`r format(Sys.time(), '%d.%m.%Y')`"
subtitle: 'Group XX: Henrik Syversveen Lie, Mikal Stapnes, Oliver Byhring'
---

```{r setup, include = FALSE}
library(formatR)
showsol <- FALSE
library(knitr)
opts_chunk$set(tidy.opts = list(width.cutoff = 68), tidy = TRUE, warning = FALSE, error = FALSE, message = FALSE, echo = TRUE)
```

# a)
```{r, echo = F, eval = T}
library(GGally)
dataset <- read.table("https://www.math.ntnu.no/emner/TMA4315/2018h/jsp2.txt", 
    header = TRUE)
ggpairs(data = dataset, mapping = aes(col = gender), columns = c("social", "raven", 
    "math"), legend = 1)
```

* Comment briefly on the plot you have created
First, we see that there is a positive correlation between the `raven` (test score) and `math` variable. This is as expected. Furthermore, we see that girls perform somewhat better in the math test than boys. Also, there is no evident correlation between social class and test scores, which is somewhat surprising.


```{r, echo = F, eval = T}
fit = lm(math~raven+gender, data = dataset)
summary(fit)


```
* Explain what the different parts of this model are called.
* Comment briefly on the parameter estimates you have found.
* What are we investigating with this model?


# b)




# c)




# d)




# e)

